Predicting Solid State Material Platforms for Quantum Technologies

### 2022-01-31 paper_v2.tex
Updates from Overleaf

%MB copied from above  
%Ferrenti 
Considering the materials present in the labeled data sets of the first two approaches more closely, we find that carbon in a diamond-like structure is classified as a suitable candidate in the Ferrenti approach, in good agreement with experimental observations (see Table~\ref{tab:qt-materials}) and hence also with the candidates that were manually defined as suitable in the empirical approach.   Interestingly, in the labeled data from the Ferrenti approach, we find that carbon in two-dimensional graphite-like structures are marked as suitable as well. All structures of silicon are also labeled as suitable together with one entry of silicon carbide. Note that this is the $3$C polytype of SiC, meaning that the most well-established quantum compatible SiC polytype, $4$H-SiC, is missing. 
% from the training set. 
Among other potentially suitable candidates we find ZnS, ZnSe, ZnO and ZnTe present in the training data.  
%Extended Ferrenti 
In the training set of the extended Ferrenti approach, we find a single entry for each of SiC, Si, GaN, ZnS, GaP, AlAs and AlP, carbon in both diamond- and graphite-like structures, and AlN in three different configurations. The labeled data set includes a larger variety of materials that are known to be quantum compatible as compared to the Ferrenti approach due to admitting more elements in the initial selection process. However, since we also included a more stringent band gap restriction of $1.5$~eV (to reduce the amount of materials in the training set), we find a more sparse representation of each known chemical formula present in the data,

@@

In order to reduce the dimensionality while preserving most of the variance in the data,
%In order to deduce how the different material features are weighted by the four machine learning models when classifying materials into suitable and unsuitable candidates for QT,
we perform a principal component analysis (PCA) for each of the three approaches.
First, the eigenvalues of the covariance matrices of the three training sets are identified in order of decreasing value,  resulting in three distinct PCA models (as opposed to the case of Fig.~\ref{fig:2dscatterplotpca} where all panels are based on the same model). 
Next, we apply the machine learning models. 
% 
In the evaluation of the machine learning models for the three different approaches we apply a $5\times 5$ stratified cross-validation when searching for the optimal hyperparameter combinations. 
%
Figure~\ref{fig:PComponents} visualizes different parameters for the 
%$25$ 
most important principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches.  
%The extracted principal components after applying the gradient boosting model for the training sets from the Ferrenti and empirical approaches are shown in Figure~\ref{fig:PComponents} in decreasing order of eigenvalue size. 
Further details on the evaluation of the machine learning models and the results from the principal component analyses for all four machine learning models are shown in the Supplementary Material at \cite{supplementary}.

Fig.~\ref{fig:PComponents}(a) displays the principal components as they are sorted in descending order by the explained variance. For the Ferrenti approach (Fig.~\ref{fig:PComponents}a) to reach the $95 \ \%$ accumulated explained variance, a total of $144$ principal components must be included. For the extended Ferrenti approach, we find that gradient boosting performs optimally at $93$ principal components with a balanced accuracy score of $0.85$ (not shown). In the case of the empirical approach and in contrast to the Ferrenti based approaches, the decision trees and random forest models exhibit the best performance for just a few principal components, and experience a considerable degree of overfitting 
%for larger principal component values. 
when involving more principal components.

Gradient boosting, in contrast to the case for the two Ferrenti based approaches, also experiences the best performance for just a few principal components. 
\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-PC-v2.tex}
        \label{fig:01-fi-e}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-GB-v2.tex}
        \label{fig:01-fi-d}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-PC-v2.tex}
        \label{fig:03-fi-e}
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-GB-v2.tex}
        \label{fig:03-fi-d}
        \subcaption{}
    \end{subfigure}
    \caption{Visualizing different parameters for the $25$ most principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches. 
    }
    \label{fig:PComponents}
\end{figure}

@@

We first consider the machine learning classification of the test set derived in the Ferrenti approach. 
From the labeled data set of $23.623$ materials, the four methods consistently predict at least $11.000$ materials as promising candidates in the case of the Ferrenti approach. All four machine learning methods agree on a total of $6804$ suitable candidates, however, many of these materials are predicted with an incidence similar to that of a coin flip. If we were to raise the minimum confidence cut-off for a prediction to (e.g.) $0.75$ instead of $0.5$, the above machine learning algorithms would only agree on $1784$ suitable candidates. 
We find that all four machine learning methods admit almost all materials with a chemical formula matching the known suitable candidates (see Table~\ref{tab:qt-materials}) that were present in the labeled data. 
This can allow materials with unfortunate structures to be labeled as suitable candidates. %by all of our machine learning methods. 
For example, the machine learning methods do not perpetuate the band gap restriction from the training set definition. 
Based on the restrictions we defined initially, we would expect that all materials with band gaps lower than $0.5$~eV would be eliminated 
%for a cut-off set to above $0.5$ 
after applying the machine learning algorithms.  
Although this is expected behavior based on the data labeling process, the trend is not carried over to the predicted data. We find that many entries with band gaps lower than $0.5$~eV are predicted as both suitable and unsuitable candidates by all four machine learning methods employed herein, 
indicating that the exact cut-off value we set for the band gap ($0.5$~eV for the Ferrenti approach) is not recognized by the machine learning algorithms --- despite the principal component analysis revealing that the band gap is an important feature for the machine learning classification in the Ferrenti approach.  
On the other hand, due to the known underestimation of band gaps by the PBE functional, the band gaps could in reality be larger for many of the materials.

@@

%histograms 
\begin{figure}[ht]
    \begin{subfigure}[b]{1\textwidth}
    \centering
    \input{figures/histograms/new/legend.tex}
  \end{subfigure}
  \par\bigskip
\begin{subfigure}[b]{0.45\textwidth}
    \input{figures/histograms/new/MP_Eg_count_075.tex}
    \subcaption{}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
    \input{figures/histograms/new/IonProperty_max ionic char_count_075.tex}
    \subcaption{}
\end{subfigure}%

@@

Interestingly, none of the nine materials predicted by the empirical approach with $0.85$ confidence, nor of the six materials predicted by all approaches to $0.75$ confidence, are elemental. At most three different elements are present, but the emphasis is clearly on binary compounds. This resonates strongly with the observations from the principal component analysis: an optimal degree of crystalline order likely exists for a material to manifest QT compatible properties, but retaining some variations throughout the crystal in (e.g.) bond length and symmetry.  Indeed, this is 
%in contrast 
contrary to the expectations that guided the formulation of the test, training and validation sets in the Ferrenti and extended Ferrenti approaches. Where most previous works have highlighted features such as band gap, polarity and ionic character as vital for a semiconductor to be a good quantum host, our results instead reveal that local variations in the crystal structure related to symmetry and bond angles, length and orientation should be considered 
%as well. 
instead.

@@

JARVIS-DFT \cite{Choudhary2020} is an open-source database based on the VASP software and consists of roughly $40.000$ three-dimensional materials using the vdW-DF-OptB88 (OPT) functional \cite{Thonhauser2007, Klimes2011}. Structures included in the database are originally taken from the Materials 
%mhj: ref to materials project is missing
Project, and then re-optimized using the OPT functional. Finally, the combination of the OPT and modified Becke-Johnson (mBJ) functionals is used to obtain a representative band gap for each structure \cite{Choudhary2018a}. Data extraction from JARVIS-DFT was done in January of $2021$, were we utilized the version made available 30.04.2021 at \url{figshare.com}.

@@

Pymatgen, a robust, open-source Python library for materials analysis \cite{pymatgen}, was also employed.

>>>>

%MB copied from above  
%Ferrenti 
Considering the materials present in the labeled data sets of the first two approaches more closely, we find that carbon in a diamond-like structure is classified as a suitable candidate in the Ferrenti approach, in good agreement with experimental observations (see Table~\ref{tab:qt-materials}) and hence also with the candidates that were manually defined as suitable in the empirical approach.   Interestingly, in the labeled data from the Ferrenti approach, we find that carbon in two-dimensional graphite-like structures are marked as suitable as well. All structures of silicon are also labeled as suitable together with one entry of silicon carbide. Note that this is the $3$C polytype of SiC, meaning that the most well-established quantum compatible SiC polytype, $4$H-SiC, is missing. 
% from the training set. 
Among other potentially suitable candidates we find ZnS, ZnSe, ZnO and ZnTe present in the training data.  
%Extended Ferrenti 
In the training set of the extended Ferrenti approach, we find a single entry for each of SiC, Si, GaN, ZnS, GaP, AlAs and AlP, carbon in both diamond- and graphite-like structures, and AlN in three different configurations. The labeled data set includes a larger variety of materials that are known to be quantum compatible as compared to the Ferrenti approach due to admitting more elements in the initial selection process. However, since we also included a more stringent band gap restriction of $1.5$~eV (to reduce the amount of materials in the training set), we find a more sparse representation of each known chemical formula present in the training set.

@@

In order to reduce the dimensionality while preserving most of the variance in the data,
%In order to deduce how the different material features are weighted by the four machine learning models when classifying materials into suitable and unsuitable candidates for QT,
we perform a principal component analysis (PCA) for each of the three approaches.
First, the eigenvalues of the covariance matrices of the three training sets are identified in order of decreasing value,  resulting in three distinct PCA models (as opposed to the case of Fig.~\ref{fig:2dscatterplotpca} where all panels are based on the same model).

Next, we apply the machine learning models. In the evaluation of the machine learning models for the three different approaches we apply a $5\times 5$ stratified cross-validation when searching for the optimal hyperparameter combinations. We note that all four machine learning models have high evaluation metrics for the optimal hyperparameters. Further details on the evaluation of the machine learning models and the results from the principal component analyses for all four machine learning models are shown in the Supplementary Material at \cite{supplementary}.

%OH: Downsides. Could also be moved to methods?
We note that there are several potential pitfalls of using the chosen dimensionality reduction and machine learning techniques in this paper. PCA is a trivial algorithm that only accounts for linear correlation between features, where other algorithms, such as Kernel PCA, can facilitate non-linear dimensionality reduction but with higher computational demand. Additionally, by using a trivial PCA, we make an assumption of orthogonality, since the eigenvectors are orthogonal to each other. However, there might exist an non-orthogonal basis which captures the non-orthogonal data in the dataset better.

%OH: Øyvind og Seb: 
%Comment on pitfalls of ML metods
%Comment on pitfalls on data size.

Figure~\ref{fig:PComponents} visualizes different parameters for the most important principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches. This differs from Figure \ref{fig:2dscatterplotpca}, where we visualize the two most important eigenvectors for all approaches that originates from the same covariance matrix.
%Fig.~\ref{fig:PComponents}(a) displays the principal components as they are sorted in descending order by the explained variance. 
For the Ferrenti approach (Fig.~\ref{fig:PComponents}a) to reach the $95 \ \%$ accumulated explained variance, a total of $144$ principal components must be included. For the extended Ferrenti approach, we find that gradient boosting performs optimally at $93$.

In the case of the empirical approach and in contrast to the Ferrenti based approaches, the decision trees and random forest models exhibit the best performance for just a few principal components, and experience a considerable degree of overfitting 
%for larger principal component values. 
when involving more principal components. Gradient boosting, in contrast to the case for the two Ferrenti based approaches, also experiences the best performance for just a few principal components. 
\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-PC-v2.tex}
        \label{fig:01-fi-e}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-GB-v2.tex}
        \label{fig:01-fi-d}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-PC-v2.tex}
        \label{fig:03-fi-e}
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-GB-v2.tex}
        \label{fig:03-fi-d}
        \subcaption{}
    \end{subfigure}
    \caption{Visualizing different parameters for the $25$ most principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches. 
    }
    \label{fig:PComponents}
\end{figure}

@@

We first consider the machine learning classification of the test set derived in the Ferrenti approach. 
From the labeled data set of $23.623$ materials, the four methods consistently predict at least $11.000$ materials as promising candidates in the case of the Ferrenti approach. All four machine learning methods agree on a total of $6804$ suitable candidates, however, many of these materials are predicted with an incidence similar to that of a coin flip. If we were to raise the minimum confidence cut-off for a prediction to e.g. $0.75$ instead of $0.5$, the above machine learning algorithms would only agree on $1784$ suitable candidates. 
We find that all four machine learning methods admit almost all materials with a chemical formula matching the known suitable candidates (see Table~\ref{tab:qt-materials}) that were present in the labeled data. 
This can allow materials with unfortunate structures to be labeled as suitable candidates. %by all of our machine learning methods. 
For example, the machine learning methods do not perpetuate the band gap restriction from the training set definition. 
Based on the restrictions we defined initially, we would expect that all materials with band gaps lower than $0.5$~eV would be eliminated 
%for a cut-off set to above $0.5$ 
after applying the machine learning algorithms.  
Although this is expected behavior based on the data labeling process, the trend is not carried over to the predicted data. We find that many entries with band gaps lower than $0.5$~eV are predicted as both suitable and unsuitable candidates by all four machine learning methods employed herein, 
indicating that the exact cut-off value we set for the band gap ($0.5$~eV for the Ferrenti approach) is not recognized by the machine learning algorithms --- despite the principal component analysis revealing that the band gap is an important feature for the machine learning classification in the Ferrenti approach.  
On the other hand, due to the known underestimation of band gaps by the PBE functional, the band gaps could in reality be larger for many of the materials.

@@

%histograms 
\begin{figure}[ht]
    \begin{subfigure}[b]{1\textwidth}
    \centering
    \input{figures/histograms/new/LegendHistPaper.tex}
  \end{subfigure}
  \par\bigskip
\begin{subfigure}[b]{0.45\textwidth}
    \input{figures/histograms/new/MP_Eg_count_075.tex}
    \subcaption{}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
    \input{figures/histograms/new/IonProperty_max ionic char_count_075.tex}
    \subcaption{}
\end{subfigure}%

@@

Interestingly, none of the nine materials predicted by the empirical approach with $0.85$ confidence, nor of the six materials predicted by all approaches to $0.75$ confidence, are elemental. At most three different elements are present, but the emphasis is clearly on binary compounds. This resonates strongly with the observations from the principal component analysis: an optimal degree of crystalline order likely exists for a material to manifest QT compatible properties, but retaining some variations throughout the crystal in e.g. bond length and symmetry.  Indeed, this is 
%in contrast 
contrary to the expectations that guided the formulation of the test, training and validation sets in the Ferrenti and extended Ferrenti approaches. Where most previous works have highlighted features such as band gap, polarity and ionic character as vital for a semiconductor to be a good quantum host, our results instead reveal that local variations in the crystal structure related to symmetry and bond angles, length and orientation should be considered 
%as well. 
instead.

@@

JARVIS-DFT \cite{Choudhary2020} is an open-source database based on the VASP software and consists of roughly $40.000$ three-dimensional materials using the vdW-DF-OptB88 (OPT) functional \cite{Thonhauser2007, Klimes2011}. Structures included in the database are originally taken from the Materials 
%mhj: ref to materials project is missing
Project \cite{Jain2013, Jain2018}, and then re-optimized using the OPT functional. Finally, the combination of the OPT and modified Becke-Johnson (mBJ) functionals \cite{Tran2009} is used to obtain a representative band gap for each structure \cite{Choudhary2018a}. Data extraction from JARVIS-DFT was done in January of $2021$, were we utilized the version made available 30.04.2021 at \url{figshare.com}.

@@

Pymatgen, a robust, open-source Python library for materials analysis \cite{pymatgen}, was also employed to extract and generate features for several of the databases mentioned above.
